---
title: "Project 3: STAT302package Tutorial"
author: "Kelsi Osorio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STAT302package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This package is a compilation of functions that run some statistical inferences and predictions that we have learned over the course of this class. The STAT302package includes functions my_t.test(), my_lm(), my_knn_cv, and my_rf_cv, as well as two sets of data my_penguins and my_gapminder from the gapminder and palmerpenguins packages. For statistical inference we have the function my_t.test that performs a one sample t-test and my_lm fits a linear model. For statistical prediction, other than my_lm, it has the my_knn_cv function that predicts an output class using covariates, and my_rf_cv, specific to the my_penguins data, which predicts output body_mass_g using covariates bill_length_mm, bill_depth_mm, and flipper_length_mm. In this vignette we will be using that data in order to demonstrate what this functions can do. 

Install STAT302package using:
```{r, eval = FALSE}
devtools::install_github("kelsiosorio/STAT302package")
```

```{r setup}
library(STAT302package)
```

# my_t.test Tutorial 




